\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-01-26 simple/latex/ms.tex
paper: simulated injection recovery results

    nu = Nuance(time, flux, error, kernel, X)
\end{lstlisting}
where \texttt{kernel} is the kernel of the Gaussian Process used to model corelated noise, and \texttt{X} the design matrix of the linear model. In \nuance{}, Gaussian Processes are created through the \texttt{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}} Python package powered by \texttt{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}}, allowing for custom kernels to be built and highly tractable computations. We can then define a set of epochs \texttt{t0} and durations \texttt{Ds} and run the \textit{linear search} with
\begin{lstlisting}[language=Python]
    import numpy as np

    t0s = time.copy()
    # a range of 20 durations
    Ds = np.linspace(0.01, 0.2, 10)
    nu.linear_search(t0s, Ds)
\end{lstlisting}
Finally, the \textit{periodic search} is run with
\begin{lstlisting}[language=Python]
    # a wide range of periods
    periods = np.linspace(0.1, 5, 2000)
    search = nu.periodic_search(periods)
\end{lstlisting}
From this \texttt{search} object can be computed the best transiting candidate parameters (\texttt{search.best}), the $\mathcal{Q}$ periodogram (\texttt{search.Q\_snr}) as well as valuable information about the transit search. The \texttt{Nuance} object also provides methods to perform transit search on light curves from multi-planetary systems, the advantage of \texttt{nuance} being that the \textit{linear search} only needs to be performed ones, and reused for the search of several transiting candidates. An extensive and maintained online documentation is provided at \href{https://nuance.readthedocs.io}{\texttt{nuance.readthedocs.io}}.

\newpage
\section{Validation using transit injection-recovery}\label{simu}
In order to validate our algorithm, we assess the transit search capabilities of \texttt{nuance} against commonly used techniques. As demonstrated in \autoref{detrending_effect}, the full-fledged modeling offered by \texttt{nuance} is not always necessary, and becomes beneficial for particular noise characteristics that are relative to the searched transit parameters. Hence, our comparison will be done on the relative parameter space $(\tau, \delta)$ described in \autoref{eq:relative_param_space}, focusing on its formulation by assuming correlated noise in the form of stellar variability (\autoref{eq:relative_params}) simulated using a Gaussian Process with an SHO kernel (\autoref{app_gp}). Ultimately, we demonstrate in this section how \texttt{nuance} compares with commonly used techniques, depending on the similarity between stellar variability and the searched transit signals.
\\\\
In what follows, we inject transits in simulated or real datasets, and employ algorithms to recover these transits with the right parameters (hence \textit{injection-recovery}). We consider the transit to be found if the absolute difference between the periods $dP$ and epochs $dT_0$ of the recovered and injected planet follow
$$dP<0.1\;\text{days} \hspace{0.5cm} \text{and} \hspace{0.5cm} dT_0<0.01\;\text{days}$$

@@

We start by comparing \texttt{naunce} to a relatively simple method on simulated light curves. As in \autoref{detrending_effect}, this simple method consist in searching for planetary transits by first modeling stellar variability using an optimal biweight filter (implemented in \texttt{wõtan}, \citealt{wotan}), subtracting it from the light-curve, and perform the search on the detrended light curve using an optimal box-least-square algorithm (\texttt{TLS}, \citealt{tls}). We name this method \wtls{}.
\\\\
The dataset consists in 10 000 light curves simulated using the model described in \autoref{app_principle_simulations}. We simulate a common periodic transit among all light curves, of period $P=1.1$ days, epoch $T_0=0.2$ days, duration $D=0.04$ days and depth $\Delta=2\%$. Each light curve consist in a 3 days observation, with an exposure time of 2 minutes, leading to $N=2160$ data points with a normal error of $0.1\%$.
\\\\
For a given pair of $(\tau, \delta)$, we simulate stellar variability using a Gaussian Process with an SHO kernel of hyperparameters defined by \autoref{eq:relative_params}, computed with respect to the injected transit parameters $D$ and $\Delta$. The same kernel is used for the search with \texttt{nuance}, an optimal choice on equal footing with the optimal biweight filter employed in the \wtls{} search. We generate 10 000 pairs of $(\tau, \delta)$ such that
$$\tau \sim \mathcal{U}(0.1, 5)\hspace{0.5cm}\text{and}\hspace{0.5cm} \delta \sim \mathcal{U}(0.1, 10)$$
with $\mathcal{U}(a, b)$ a uniform distribution of lower bound $a$ and upper bound $b$.

\vfill
\TODO
\begin{figure}[H]
    \begin{centering}
        \includegraphics[height=12cm]{../../workflows/synthetic_injection_recovery/figures/final_result.pdf}
        \caption{\TODO}
        \label{fig:simu}
    \end{centering}
\end{figure}

>>>>

    nu = Nuance(time, flux, error, kernel, X)
\end{lstlisting}
where \texttt{kernel} is the kernel of the Gaussian Process used to model corelated noise, and \texttt{X} the design matrix of the linear model. \nuance{} exploits the Gaussian Process implementation from \texttt{tinygp}\footnote{\href{https://github.com/dfm/tinygp}{https://github.com/dfm/tinygp}}, a Python package powered by \texttt{JAX}\footnote{\href{https://github.com/google/jax}{https://github.com/google/jax}} allowing for custom kernels to be built and highly tractable computations. We can then define a set of epochs \texttt{t0} and durations \texttt{Ds} and run the \textit{linear search} with
\begin{lstlisting}[language=Python]
    import numpy as np

    t0s = time.copy()
    # a range of 20 durations
    Ds = np.linspace(0.01, 0.2, 10)
    nu.linear_search(t0s, Ds)
\end{lstlisting}
Finally, the \textit{periodic search} is run with
\begin{lstlisting}[language=Python]
    # a wide range of periods
    periods = np.linspace(0.1, 5, 2000)
    search = nu.periodic_search(periods)
\end{lstlisting}
From this \texttt{search} object, the best transiting candidate parameters can be computed (\texttt{search.best}), or the $\mathcal{Q}$ periodogram retrieved (\texttt{search.Q\_snr}), together with valuable information about the transit search. The \texttt{Nuance} object also provides methods to perform transit search on light curves from multi-planetary systems, the advantage of \texttt{nuance} being that the \textit{linear search} only needs to be performed ones, and reused for the search of several transiting candidates. An extensive and maintained online documentation is provided at \href{https://nuance.readthedocs.io}{\texttt{nuance.readthedocs.io}}.

\newpage
\section{Validation with injection-recovery}\label{simu}
In order to validate our algorithm, we assess the transit search capabilities of \texttt{nuance} against commonly used techniques. As demonstrated in \autoref{detrending_effect}, the full-fledged modeling offered by \texttt{nuance} is not always necessary, and becomes beneficial for particular noise characteristics that are relative to the searched transit parameters. Hence, our comparison will be done on the relative parameter space $(\tau, \delta)$ described in \autoref{eq:relative_param_space}, focusing on its formulation by assuming correlated noise in the form of stellar variability (\autoref{eq:relative_params}) simulated using a Gaussian Process with an SHO kernel (\autoref{app_gp}). Ultimately, we demonstrate in this section how \texttt{nuance} compares with commonly used techniques, depending on the similarity between stellar variability and the searched transit signals.
\\\\
In what follows, we inject transits in simulated or real datasets, and employ algorithms to recover these transits with the right parameters (hence \textit{injection-recovery}). We consider the transit to be found if the absolute difference between the periods $dP$ and epochs $dT_0$ of the recovered and injected planets follow
$$dP<0.1\;\text{days} \hspace{0.5cm} \text{and} \hspace{0.5cm} dT_0<0.01\;\text{days}$$

@@

We start by comparing \texttt{naunce} to a relatively simple method on simulated light curves. As in \autoref{detrending_effect}, this simple method consist in searching for planetary transits by first modeling stellar variability using an optimal biweight filter (implemented in \texttt{wõtan} \citep{wotan} using a window size three times that of the injected transit duration), subtracting it from the light-curve, and then perform the search on the detrended light curve using an optimal box-least-square algorithm (\texttt{TLS}, \citealt{tls}). We name this method \wtls{}.
\\\\
The dataset consists in 10 000 light curves simulated using the model described in \autoref{app_principle_simulations}. We simulate a common periodic transit among all light curves, of period $P=1.1$ days, epoch $T_0=0.2$ days, duration $D=0.04$ days and depth $\Delta=2\%$. Each light curve consist in a 3 days observation, with an exposure time of 2 minutes, leading to $N=2160$ data points with a normal error of $0.1\%$.
\\\\
For a given pair of $(\tau, \delta)$, we simulate stellar variability using a Gaussian Process with an SHO kernel of hyperparameters defined by \autoref{eq:relative_params}, computed with respect to the injected transit parameters $D$ and $\Delta$. The same kernel is used for the search with \texttt{nuance}, an optimal choice on equal footing with the optimal biweight filter employed in the \wtls{} search. We generate 10 000 pairs of $(\tau, \delta)$ such that
$$\tau \sim \mathcal{U}(0.1, 40)\hspace{0.5cm}\text{and}\hspace{0.5cm} \delta \sim \mathcal{U}(0.1, 100)$$
where $\mathcal{U}(a, b)$ denotes a uniform distribution of lower bound $a$ and upper bound $b$.
\\\\
The results of these injection-recovery are shown in \autoref{fig:simu} and highlight particularly well the benefit of \nuance{} against \wtls{} on transits corresponding to $\delta > 3$ and $\tau > 2$, i.e. with relatively low transit depth compared to the stellar variability amplitude, and a relatively small duration compared to the stellar variability period. On average, searching a single light curve with \nuance{} took 9 seconds, compared to 8 seconds with \wtls{}. Beyond its increased sensitivity, this metric demonstrates the tractability of \nuance{}, studied in more details in \autoref{}.
\\\\
About these results, we note these transit searches were done in a particularly optimal setup, on light curves not all physically realistic, hence demonstrating the performance of \nuance{} on a theoretical basis only. In the next subsection, we perform transits injection-recovery on a real dataset, using more realistic tools and on light curves featuring known planetary transits, hence validating \nuance{} on a realistic multi-planetary dataset.

\begin{figure}[H]
    \begin{centering}
        \includegraphics[height=12cm]{../../workflows/synthetic_injection_recovery/figures/final_result.pdf}
        \caption{\TODO}
        \label{fig:simu}
    \end{centering}
\end{figure}
